The evolution of the transcriptome of mammalian placentas

### 2016-11-14 manuscript/manuscript.Rnw
add Priya
move core placenta distribution shape to paper

\title{The evolution of the transcriptome of mammalian placentas}%
%
\author[affigb]{Don L. Armstrong\fnref{n1}}
\ead{don@donarmstrong.com}
\author[afflon]{Michael R. McGowen\fnref{n1}}
\ead{1mmcgowen1@gmail.com}
\author[affigb]{Amy Weckle}
\ead{aweckle@illinois.edu}
\author[affwayne]{Jason Caravas}
\ead{jacaravas@gmail.com}
\author[affmich]{Dalen Agnew}
\ead{agnewd@dcpah.msu.edu}
\author[affuscd]{Kurt Benirschke}
\ead{kbenirschke@ucsd.edu}
\author[affapetrust]{Sue Savage-Rumbaugh}
\ead{suerumbaugh@gmail.com}
\author[affhaifa]{Eviatar Nevo}
\ead{nevo@research.haifa.ac.il}
\author[affnih,affwaynepath]{Chong J Kim}
\ead{cjkim@med.wayne.edu}
\author[affyale]{Günter P. Wagner}
\ead{gunter.wagner@yale.edu}
\author[affnih]{Roberto Romero}
\ead{prbchiefstaff@med.wayne.edu}
\author[affigb,affapetrust,affmip]{Derek E. Wildman\corref{cor1}}
\ead{wildmand@illinois.edu}
\cortext[cor1]{Corresponding author}
\address[affigb]{Institute for Genomic Biology, University of Illinois at Urbana-Champaign, Urbana, IL, USA}
\address[afflon]{School of Biological and Chemical Sciences, Queen Mary, University of London, London, UK}
\address[affwayne]{Center for Molecular Medicine and Genetics, Wayne State University School of Medicine, Detroit, MI, USA}
\address[affmich]{Diagnostic Center for Population and Animal Health, College of Veterinary Medicine, Michigan State University, East Lansing, MI, USA}
\address[affuscd]{Department of Pathology, University of California San Diego, La Jolla, CA, USA}
\address[affapetrust]{Great Ape Trust/Bonobo Hope Sanctuary, Des Moines, IA, USA}
\address[affnih]{Perinatology Research Branch, Eunice Kennedy Shriver National Institute of Child Health and Human Development (NICHD), National Institutes of Health (NIH), Detroit, MI, USA}
\address[affwaynepath]{Department of Pathology, Wayne State University School of Medicine, Detroit, MI, USA}
\address[affyale]{Department of Ecology and Evolutionary Biology and Yale Systems Biology Institute, Yale University, New Haven, CT, USA}
\address[affmip]{Department of Molecular and Integrative Physiology, University of Illinois at Urbana-Champaign, Urbana, IL, USA}
\address[affhaifa]{The Institute of Evolution, University of Haifa, Haifa, Israel}

@@

<<core.placenta.transcriptome.distribution,results="hide",echo=FALSE,message=FALSE>>=
num.species <- star.logs.wide[,length(unique(species))]
calculate.distribution.shape <- 
    function(min.vector) {
        min.vector <- sort(min.vector,decreasing=TRUE)
        temp <- data.table(FPKM=min.vector,num_genes=1:length(min.vector))
        temp <- temp[order(-num_genes)][!duplicated(FPKM)]
        return(temp)
    }

@@

cpt.distribution.shape <- 
    lapply(c("all_expression",
             "all_but_one_expression",
             "all_but_two_expression",
             "all_but_three_expression",
             "all_but_four_expression"),
           function(x){
               calculate.distribution.shape(cptg.nh[,x,with=FALSE][[1]])[,`Minimum Expression in`:=x]
           })
fix.expression.names <- function(x){
    x <- gsub("^all_expression$",
              paste0(num.species," of ",num.species," species"),
              x
              )
    nc.levels <- paste0(num.species," of ",num.species," species")
    ## number coutner
    nc <- c("one"=1,"two"=2,"three"=3,"four"=4)
    for (a in names(nc)) {
        x <- gsub(paste0("^all_but_",a,"_expression$"),
              paste0(num.species-nc[a]," of ",num.species," species"),
              x
              )
        nc.levels <-
            c(nc.levels,
              paste0(num.species-nc[a]," of ",num.species," species"))
    }
    return(factor(x,levels=nc.levels))
}

@@

cpt.distribution.shape <- 
    cpt.distribution.shape[,`Minimum Expression in`:=
                                fix.expression.names(`Minimum Expression in`)]

setnames(cpt.distribution.shape,
         "FPKM",
         "Expression (FPKM)")
@

The
\Sexpr{core.placenta.transcriptome.long[,length(unique(human_name))]}
genes comprising the core placenta transcriptome are shown in
\cref{fig:core_placenta_transcriptome} ordered by their median
transcription level in \ac{FPKM}. We considered a gene to be core to
the placenta transcriptome if a 1:1 ortholog of that gene was
expressed with $\text{FPKM} \ge 10$ in all species studied and was
not a human housekeeping gene according to
\citet{Eisenberg.ea2013:Humanhousekeepinggenesrevisited}.
Supplemental \cref{fig:core_placenta_transcriptome_shape} shows the number of genes
which would be included in the core placenta set if the \ac{FPKM}
cutoff was set differently; for example, choosing an \ac{FPKM} cutoff of 1 yields %
\Sexpr{cpt.distribution.shape[`Expression (FPKM)`>1 & 
`Minimum Expression in` == '14 of 14 species',max(num_genes)]} %
genes in the core transcriptome. %
In contrast to the shape of the number of genes included in the core
placenta transcriptome, the set of \human housekeeping genes expressed
all placentas is flatter (Supplemental
\cref{fig:core_housekeeping_transcriptome_shape}), and includes more
of the total fraction of genes. However, there is still a large
difference between the set of housekeeping genes which are expressed
in \human and those which are expressed in all placentas.

@@

>>>>

\title{The evolution of the transcriptome of mammalian placentas}%
%
\author[affigb]{Don L. Armstrong\fnref{n1}}
\ead{don@donarmstrong.com}
\author[afflon]{Michael R. McGowen\fnref{n1}}
\ead{1mmcgowen1@gmail.com}
\author[affigb]{Amy Weckle}
\ead{aweckle@illinois.edu}
\author[affigb]{Priyadarshini Pantham}
\ead{pantham@illinois.edu}
\author[affwayne]{Jason Caravas}
\ead{jacaravas@gmail.com}
\author[affmich]{Dalen Agnew}
\ead{agnewd@dcpah.msu.edu}
\author[affuscd]{Kurt Benirschke}
\ead{kbenirschke@ucsd.edu}
\author[affapetrust]{Sue Savage-Rumbaugh}
\ead{suerumbaugh@gmail.com}
\author[affhaifa]{Eviatar Nevo}
\ead{nevo@research.haifa.ac.il}
\author[affnih,affwaynepath]{Chong J Kim}
\ead{cjkim@med.wayne.edu}
\author[affyale]{Günter P. Wagner}
\ead{gunter.wagner@yale.edu}
\author[affnih]{Roberto Romero}
\ead{prbchiefstaff@med.wayne.edu}
\author[affigb,affapetrust,affmip]{Derek E. Wildman\corref{cor1}}
\ead{wildmand@illinois.edu}
\cortext[cor1]{Corresponding author}
\address[affigb]{Institute for Genomic Biology, University of Illinois at Urbana-Champaign, Urbana, IL, USA}
\address[afflon]{School of Biological and Chemical Sciences, Queen Mary, University of London, London, UK}
\address[affwayne]{Center for Molecular Medicine and Genetics, Wayne State University School of Medicine, Detroit, MI, USA}
\address[affmich]{Diagnostic Center for Population and Animal Health, College of Veterinary Medicine, Michigan State University, East Lansing, MI, USA}
\address[affuscd]{Department of Pathology, University of California San Diego, La Jolla, CA, USA}
\address[affapetrust]{Great Ape Trust/Bonobo Hope Sanctuary, Des Moines, IA, USA}
\address[affnih]{Perinatology Research Branch, Eunice Kennedy Shriver National Institute of Child Health and Human Development (NICHD), National Institutes of Health (NIH), Detroit, MI, USA}
\address[affwaynepath]{Department of Pathology, Wayne State University School of Medicine, Detroit, MI, USA}
\address[affyale]{Department of Ecology and Evolutionary Biology and Yale Systems Biology Institute, Yale University, New Haven, CT, USA}
\address[affmip]{Department of Molecular and Integrative Physiology, University of Illinois at Urbana-Champaign, Urbana, IL, USA}
\address[affhaifa]{The Institute of Evolution, University of Haifa, Haifa, Israel}

@@

\begin{figure}
<<core.placenta.transcriptome.distribution,fig.width=5,fig.height=5,out.height="0.8\\textheight",echo=FALSE,message=FALSE>>=
num.species <- star.logs.wide[,length(unique(species))]
calculate.distribution.shape <- 
    function(min.vector) {
        min.vector <- sort(min.vector,decreasing=TRUE)
        temp <- data.table(FPKM=min.vector,num_genes=1:length(min.vector))
        temp <- temp[order(-num_genes)][!duplicated(FPKM)]
        return(temp)
    }

@@

cpt.distribution.shape <- 
    lapply(c("all_expression",
             "all_but_one_expression",
             "all_but_two_expression",
             "all_but_three_expression",
             "all_but_four_expression"),
           function(x){
               calculate.distribution.shape(cptg.nh[,x,with=FALSE][[1]])[,`Min. Exp. in`:=x]
           })
fix.expression.names <- function(x){
    x <- gsub("^all_expression$",
              paste0(num.species," of ",num.species),
              x
              )
    nc.levels <- paste0(num.species," of ",num.species)
    ## number coutner
    nc <- c("one"=1,"two"=2,"three"=3,"four"=4)
    for (a in names(nc)) {
        x <- gsub(paste0("^all_but_",a,"_expression$"),
              paste0(num.species-nc[a]," of ",num.species),
              x
              )
        nc.levels <-
            c(nc.levels,
              paste0(num.species-nc[a]," of ",num.species))
    }
    return(factor(x,levels=nc.levels))
}

@@

cpt.distribution.shape <- 
    cpt.distribution.shape[,`Min. Exp. in`:=
                                fix.expression.names(`Min. Exp. in`)]

setnames(cpt.distribution.shape,
         "FPKM",
         "Expression (FPKM)")

print(ggplot(cpt.distribution.shape[`Expression (FPKM)` > 0.1,],
             aes(x=`Expression (FPKM)`,color=`Min. Exp. in`,y=num_genes))
      + geom_line(lwd=1)
      + xlab("Minimum Expression cutoff (FPKM)")
      + ylab("Number of Genes Meeting FPKM Cutoff in N species")
      + scale_y_log10(breaks = trans_breaks("log10", function(x) 10^x),
                      labels = trans_format("log10", math_format(10^.x)))
      + scale_x_log10(breaks = trans_breaks("log10", function(x) 10^x),
                      labels = trans_format("log10", math_format(10^.x)))
      + geom_vline(xintercept=10,linetype="dotted")
      + geom_segment(mapping=aes(y=V1,
                         yend=V1,
                         xend=10,
                         x=300,
                         color=`Min. Exp. in`),
                     data=cpt.distribution.shape[`Expression (FPKM)` > 10,
                                              max(num_genes),`Min. Exp. in`],
                     arrow=arrow(length=unit(0.5,"cm")),
                     show.legend=FALSE
                     )
      + geom_text(mapping=aes(x=300,
                      y=V1,
                      label=paste0(V1," genes"),override=TRUE),
                  data=cpt.distribution.shape[`Expression (FPKM)` > 10,
                      max(num_genes),`Min. Exp. in`],
                  show.legend=FALSE,
                  hjust=0
                  )
      + guides(color=guide_legend(title=paste0("Min. Expression in\nn of ",num.species," Species",
                                  keywidth=unit(3,"mm"),
                                  keyheight=unit(3,"mm")
                                  ))
      + theme_bw()
      + theme(legend.position=c(1,1),
              legend.margin=unit(0,"mm"),
              legend.justification=c(1,1),
              plot.margin=unit(c(0,0,0,0),"mm")
              )
      )
@ 
\caption{Distribution shape of placenta core genes. The vertical axis
  shows the number of 1:1 orthologous genes included in the core
  placenta transcriptome if %
  \Sexpr{cpt.distribution.shape[,levels(`Min. Exp. in`)[1]]}
  %
  (or %
  \Sexpr{cpt.distribution.shape[,levels(`Min. Exp. in`)[2]]},
  % 
  etc.) have a minimum expression in FPKM equal to the horizontal axis.
  For example, if the cutoff is set to 10 (vertical black line), 
  % 
  \Sexpr{cpt.distribution.shape[`Expression (FPKM)` > 10 
    & `Min. Exp. in` == '14 of 14',max(num_genes)]}
  %
  genes would be included if the FPKM of a set of 1:1 orthologs of all 14 species must be at
  least 10, %
  \Sexpr{cpt.distribution.shape[`Expression (FPKM)` > 10 
    & `Min. Exp. in` == '13 of 14',max(num_genes)]}
  %
  1:1 orthologs would be included if 13 of 14 species have an
   with an FPKM of at least 10, and 
  \Sexpr{cpt.distribution.shape[`Expression (FPKM)` > 10 
    & `Min. Exp. in` == '10 of 14',max(num_genes)]}
  genes would be included if 10 of 14 species
  must have an FPKM of at least 10.}
\label{fig:core_placenta_transcriptome_shape}
\end{figure}

The
\Sexpr{core.placenta.transcriptome.long[,length(unique(human_name))]}
genes comprising the core placenta transcriptome are shown in
\cref{fig:core_placenta_transcriptome} ordered by their median
transcription level in \ac{FPKM}. We considered a gene to be core to
the placenta transcriptome if a 1:1 ortholog of that gene was
expressed with $\text{FPKM} \ge 10$ in all species studied and was
not a human housekeeping gene according to
\citet{Eisenberg.ea2013:Humanhousekeepinggenesrevisited}.
\Cref{fig:core_placenta_transcriptome_shape} shows the number of genes
which would be included in the core placenta set if the \ac{FPKM}
cutoff was set differently; for example, choosing an \ac{FPKM} cutoff of 1 yields %
\Sexpr{cpt.distribution.shape[`Expression (FPKM)`>1 & 
`Min. Exp. in` == '14 of 14',max(num_genes)]} %
genes in the core transcriptome. %
In contrast to the shape of the number of genes included in the core
placenta transcriptome, the set of \human housekeeping genes expressed
all placentas is flatter (Supplemental
\cref{fig:core_housekeeping_transcriptome_shape}), and includes more
of the total fraction of genes. However, there is still a large
difference between the set of housekeeping genes which are expressed
in \human and those which are expressed in all placentas.

@@

  Priyadarshini Pantham wrote the paper and performed gene enrichment
  analyses.
